---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%"
)
```

# GrowthCurveME

<!-- badges: start -->
<!-- badges: end -->

In-vitro growth assays often exhibit high levels of technical and experimental 
variability such as, differences in seeding density, well-to-well variation, 
and experimental replicate variation, which can impact the accuracy of growth 
estimation and modeling. The goal of GrowthCurveME is to address these aspects 
of variability by allowing users to easily fit linear and non-linear 
mixed-effects regression models to account for clustering 
(such as at the well-to-well level or experimental level) and 
longitudinal repeated measurements. GrowthCurveME makes it easy to derive key
growth metrics and model statistics, perform model diagnostics, and generate
plots and tables for reporting your data.

## Installation

You can install the development version of GrowthCurveME 
from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cancermodels-org/GrowthCurveME")
```

## Getting Started

First load the package and import a sample dataset. Here in this example we 
have sample confluency based growth data. In this experiment, there are 
multiple "clusters" or well replicates. Although they all follow a general 
growth pattern, each of the well replicates start at different growth 
measurements. 

```{r example-data}
# Load the GrowthCurveME package
library(GrowthCurveME)

# Load the ggplot2 and viridis package for graphing
library(ggplot2)
library(viridis)

# Load the exp_mixed_data dataset from GrowthCurveME package
data("exp_mixed_data")

# Inspect the data
str(exp_mixed_data)

# Create simple scatterplot of exp_mixed_data
ggplot(exp_mixed_data,
       aes(x = time, y = growth_metric, color = cluster)) +
  geom_point(size = 2) +
  scale_color_viridis(discrete = TRUE) +
  ylim(0, 25) +
  ggtitle("Example Growth vs Time Data") +
  theme_classic()

```

GrowthCurveME allows one to account for the well-to-well variation by fitting
a mixed-effects model with the growth_curve_model_fit() function. In this 
case, the data appears to follow an exponential curve, so one can set the 
function_type to "exponential". GrowthCurveME is also able to fit "linear", 
"logistic", and "gompertz" shaped functions to data. By default, the 
growth_curve_model_fit() function will return a list object that can be used 
with other function in the GrowthCurveME package. We can also fit a 
least-squares model for comparison, which will NOT account for clustering.  

```{r curve-fit, message=FALSE}
# Fit an mixed-effect model and return a summary list output
exp_mix_model_summary <- growth_curve_model_fit(
  data_frame = exp_mixed_data,
  function_type = "exponential",
  model_type = "mixed",
  time_unit = "hours")

# Fit a least-squares model for comparison
exp_ls_model_summary <- growth_curve_model_fit(
  data_frame = exp_mixed_data,
  function_type = "exponential",
  model_type = "least-squares",
  time_unit = "hours")

```

Users can produce a summary table of the model results and derived 
growth metrics for reporting. We can see in this example that the confidence 
interval for the rate and doubling time estimates are narrower than the
least-squares regression model after accounting for the inherit well-to-well
variation. Model complexity metrics such as Akaike information criterion (AIC),
Bayesian information criterion (BIC), and Log likelihood show that the 
mixed-effects is better suited for the data with clear hierarchical structures
and longitudinal measurements. 

```{r table}
# Summarize mixed-effects model
growth_model_summary_table(
  growth_model_summary_list = exp_mix_model_summary
)
# Summarize least-squares model
growth_model_summary_table(
  growth_model_summary_list = exp_ls_model_summary
)
```

Users can inspect the results of the model by producing different types of 
scatter plots with ggplot2 customization. 

```{r plots}
# Plot growth vs time data with cluster level predictions 
growth_vs_time_plot(
  growth_model_summary_list = exp_mix_model_summary,
  plot_type = 2,
  growth_metric_name = "Confluency",
  time_name = "Time (hours)",
  cluster_name = "Well",
  plot_title = "Confluency vs Time",
  y_limits = c(0, 25)
)

# Plot growth vs time data with cluster level predictions faceted by cluster
growth_vs_time_plot(
  growth_model_summary_list = exp_mix_model_summary,
  plot_type = 3,
  growth_metric_name = "Confluency",
  time_name = "Time (hours)",
  cluster_name = "Well",
  plot_title = "Confluency vs Time with Cluster-Level Predictions",
  y_limits = c(0, 25)
)

```

User can also perform model diagnostics for mixed-effect (ME) and 
least-squares (LS) models.
```{r diagnostics}
# Check residual diagnostics of model
growth_model_residual_plots(
  growth_model_summary_list = exp_mix_model_summary,
  residual_type = "cluster",
  weighted = TRUE
)

```

Users can also generate bootstrapped confidence interval plots by specifying 
'boostrap_time = TRUE' in the growth_curve_model_fit() and specifying 
'plot_type = 4' in the growth_vs_time_plot() function.
```{r boot_ci, message=FALSE, warning=FALSE}
# Fit an mixed-effect model and include boostrap estiamtes
exp_mix_model_summary <- growth_curve_model_fit(
  data_frame = exp_mixed_data,
  function_type = "exponential",
  model_type = "mixed",
  time_unit = "hours",
  bootstrap_time = TRUE,
  boot_n_sim = 10)

# Graph the boostrapped confidence intervals
growth_vs_time_plot(
  growth_model_summary_list = exp_mix_model_summary,
  plot_type = 4,
  growth_metric_name = "Confluency",
  time_name = "Time (hours)",
  cluster_name = "Well",
  plot_title = "Boostraped CI Confluency vs Time",
  y_limits = c(0, 25)
)

```


